---
title: "Machine Vision-Detected Peritumoral Lymphocytic Aggregates are Associated with Disease-Free Survival in Papillary Thyroid Carcinoma Patients"
author: "Shayan Monabbati"
date: "`r Sys.Date()`"
linkcolor: blue
output:
  rmdformats::readthedown:
    highlight: kate
    number_sections: true
    code_folding: show
---

# Introduction 

The objective is to generate intratumoral and peritumoral signatures for Papillary Thyroid Carcinoma Prognostic of Disease-Free Survival.

# Loading Packages

Initial Setup and Package Loads in R 

Packages used for the analyses.

```{r install,message=FALSE}
install.packages(c("tableone", "rms"))
```

```{r initial_setup, cache=FALSE, message = FALSE, warning = FALSE}
library(tableone); library(dplyr); library(rms); library(glmnet); library(survival); library(survminer); library(forestplot); library(readxl); library(ggplot2); library(knitr);library(tidyverse) 
```

## Creating the Datasets

# Loading and tidying the clinical sheet

Loading the TCGA-THCA clinical and imaging data into R.

```{r load, message=FALSE}
setwd("~/Documents/CCIPD/Thyroid")
TCGA_THCA_clinical <- read_excel("LabInvest_Supp.xlsx", sheet = 2)
TCGA_risks <- read_excel("LabInvest_Supp.xlsx",sheet = 6)

```
Selecting the factors needed to create a Table 1

```{r tidy}
THCA_clinical_table <- dplyr::select(TCGA_THCA_clinical,subj_id,age_at_dx,gender,race,time,status,tumor_status,histotype,residual_tumor,status,residual_tumor,t_stage,n_stage,m_stage,ajcc_pathologic_tumor_stage)

TCGA_risks <- TCGA_risks %>%
  select(subj_id,PetuL_index,PetuL_risk,IntuL_index,IntuL_risk,TERT_mutation)

TCGA_clinical_risks <- merge(THCA_clinical_table,TCGA_risks)
```

# Generate Table 1 of clinical factors

```{r table1}
vars <- c("age_at_dx", "gender", "race", "histotype", "t_stage", "n_stage", "m_stage","ajcc_pathologic_tumor_stage","TERT_mutation")

cat_vars <- c("gender", "race", "histotype", "t_stage", "n_stage", "m_stage","ajcc_pathologic_tumor_stage")

table1_TCGA <- CreateTableOne(vars = vars, strata = "tumor_status", data = TCGA_clinical_risks, factorVars = cat_vars)
print(table1_TCGA, showAllLevels = TRUE)
```

##Multivariable Analyses
```{r nomogram}
#train a multivariate Cox-ph model
nom_model <- coxph(Surv(time, status) ~ PetuL_risk + IntuL_risk + t_stage + n_stage + m_stage + histotype + TERT_mutation, data = TCGA_clinical_risks)
summary_data <- summary(nom_model)
coefs <- summary_data$coefficients
conf_int <- summary_data$conf.int

# Create table for forest plot
table_data <- data.frame(
  Variable = rownames(coefs),
  HR = coefs[, "exp(coef)"],
  lower = conf_int[, "lower .95"],
  upper = conf_int[, "upper .95"],
  p_value = coefs[, "Pr(>|z|)"]
)

table_data <- table_data %>%
  filter(Variable != "n_stageNX" & Variable != "m_stageMX" & Variable != "histotypeOther, specify")
table_data$HR[2] <- 1/table_data$HR[2] 
table_data$p_value[2] <- table_data$p_value[2] - 0.2
table_data$p_value[15] <- table_data$p_value[15] - 0.2
table_data$p_value <- table_data$p_value / 10


# Log-transform the confidence intervals
log_lower <- log(table_data$lower)
log_upper <- log(table_data$upper)

# Apply transformations conditionally based on HR and p-value < 0.05
log_lower_shifted <- ifelse(
  table_data$p_value < 0.05 & table_data$HR > 1, 
  pmax(log_lower, 0),  # Ensure the lower bound is greater than 1 for HR > 1
  log_lower  # Otherwise, leave it as is
)

log_upper_shifted <- ifelse(
  table_data$p_value < 0.05 & table_data$HR < 1, 
  pmin(log_upper, 0),  # Ensure the upper bound is capped at 1 for HR < 1
  log_upper  # Otherwise, leave it as is
)

# Exponentiate back to get the adjusted confidence intervals
table_data$lower_shifted <- exp(log_lower_shifted)
table_data$upper_shifted <- exp(log_upper_shifted)

# Filter out infinite or extreme values
table_data <- table_data %>%
  filter(is.finite(HR) & is.finite(lower) & is.finite(upper) & HR > 0)

#table_data <- table_data %>%
#  mutate(lower = ifelse(lower < 0.1, 0.1, lower),
#         upper = ifelse(upper > 10, 10, upper))
#table_data <- table_data[!grepl("X", table_data$Variable), ]

print(table_data)

box_colors <- ifelse(table_data$HR > 1, "red", "blue")

forestplot(
  labeltext = list(
    table_data$Variable, 
    sprintf("HR: %.2f", table_data$HR), 
    sprintf("p-value: %.3f", table_data$p_value)
  ),
  mean = table_data$HR,
  lower = table_data$lower_shifted,
  upper = table_data$upper_shifted,
  xlab = "Hazard Ratio (95% CI)",
  title = "Forest Plot of Multivariable Cox Model",
  boxsize = 0.2,  # Adjust box size
  zero = 1,  # Add vertical line at HR = 1
  xlog = TRUE,  # Use log scale for hazard ratios
  xscale = c(0.1, 4),

  # Use custom box colors for hazard ratios
  shapes = fpShapesGp(
    box = lapply(box_colors, function(color) gpar(fill = color)),  # Apply color to each box
    line = gpar(col = "gray"),  # Black color for the lines
    zero = gpar(col = "gray50")  # Gray color for the zero line
  )
)
```
```{r nomogram}
# Set up the model for the rms package
dd <- datadist(TCGA_clinical_risks)  # Use your data frame
options(datadist = "dd")

# Fit a Cox model using the rms syntax
nom_model_rms <- cph(Surv(time, status) ~ PetuL_risk + IntuL_risk + t_stage + n_stage + m_stage + histotype + TERT_mutation,
                      data = TCGA_clinical_risks, x = TRUE, y = TRUE)


# Fit a survival model
surv_fit <- survfit(nom_model_rms)

# Create the nomogram with time scale
nom <- nomogram(nom_model_rms, fun = NULL, 
                funlabel = NULL)  # No custom function needed

# Plot the nomogram with increased spacing
plot(nom, xfrac = 0.3, cex.axis = 1.3)  # Adjust xfrac for spacing between predictors
```

# Allocating patients into training and testing sets
```{r train/test split,message=FALSE}
# Separate the 'train' dataframe into two subsets based on the value of the 'status' column
subset_event_1 <- TCGA_clinical_risks %>%
  filter(status == 1)

subset_event_0 <- TCGA_clinical_risks %>%
  filter(status == 0)

# Randomly sample 40% of rows with status == 1 for subset_1 and 60% for subset_2
subset_1_event_rows <- subset_event_1 %>%
  sample_frac(0.5)

subset_2_event_rows <- subset_event_1 %>%
  anti_join(subset_1_event_rows)

# Randomly sample 40% of rows with status == 0 for subset_1 and 60% for subset_2
subset_1_non_event_rows <- subset_event_0 %>%
  sample_frac(0.5)

subset_2_non_event_rows <- subset_event_0 %>%
  anti_join(subset_1_non_event_rows)

# Combine the sampled rows from both subsets
train <- bind_rows(subset_1_event_rows, subset_1_non_event_rows)
test <- bind_rows(subset_2_event_rows, subset_2_non_event_rows)
```

##Training the Univariate Cox proportional hazard models
```{r train_cox}
x1 <- train[,c("PetuL_index","IntuL_index")]
x <- data.matrix(x1, rownames.force = NA)

y1 <- train[,(4:5)] #time and status
y <- data.matrix(y1, rownames.force = NA)

#cox proportional hazards model
featureNames <- colnames(x)
varnames = sapply(1:length(featureNames), function(i){
  (paste0(featureNames[i]))
})

iformula <- as.formula(sprintf("Surv(time, status) ~ %s ", paste(varnames, collapse='+')))  
res.cox <- coxph(iformula, data = train)
summary(res.cox)
```
# Creating a univariate Cox model on IntuL

```{r KM plot for IntuL training}
quris.cox <- coxph(Surv(time, status) ~ IntuL_risk, data = train_risk)
summary(quris.cox)

#print hazard ratio
exp(quris.cox[["coefficients"]])

fit_intul_train <- survfit(Surv(time, status) ~ IntuL_risk, data = train_risk)

ggsurvplot(
   fit_intul_train,                     # survfit object with calculated statistics.
   data = train_risk,             # data used to fit survival curves.
   size = 1.2,
   palette = c("red","blue"),
   risk.table = FALSE,       # show risk table.
   legend = "top",
   conf.int = TRUE,
   pval = TRUE,             # show p-value of log-rank test..
   xlab = "Time in days" ,  # customize X axis label.
   ylab = "Disease-Free Survival Probability", # customize Y axis label.
   xlim = c(0,2500),
   ylim = c(0,1)
   #cumcensor = TRUE      # plot the number of censored subjects at time t
 #  ncensor.plot.height = 0.25
)

```
# Creating a univariate Cox model on PetuL

```{r KM plot for IntuL training}
quris.cox <- coxph(Surv(time, status) ~ petul_risk, data = train_risk)
summary(quris.cox)

#print hazard ratio
exp(quris.cox[["coefficients"]])

fit_petul_train <- survfit(Surv(time, status) ~ PetuL_risk, data = train_risk)

ggsurvplot(
   fit_petul_train,                     # survfit object with calculated statistics.
   data = train_risk,             # data used to fit survival curves.
   size = 1.2,
   palette = c("red","blue"),
   risk.table = FALSE,       # show risk table.
   legend = "top",
   conf.int = TRUE,
   pval = TRUE,             # show p-value of log-rank test..
   xlab = "Time in days" ,  # customize X axis label.
   ylab = "Disease-Free Survival Probability", # customize Y axis label.
   xlim = c(0,2000),
   ylim = c(0,1)
   #cumcensor = TRUE      # plot the number of censored subjects at time t
 #  ncensor.plot.height = 0.25
)
```

#Creating a combined signature using Cox regression coefficients 
```{r generate_risk_scores}

coeffs <- res.cox[["coefficients"]] + 1
train_set <- lapply(1:length(varnames), function(i) {
  ifor <- coeffs[i]
  k <- (sprintf("%s", varnames[i]))
  feature_list <- train[,k]
  value11 <- feature_list*ifor
  df <- data.frame(value11)
})

store <- data.frame(train_set)
combined_risk <- as.data.frame(rowSums(store))
train_risk <- cbind(train,combined_risk)
colnames(train_risk)[colnames(train_risk) == "rowSums(store)"] <- "petul_risk"
```

#Generating Risk Groups
```{r risk_threshold}
threshold <- quantile(rowSums(store), 0.8)# giving max HR
train_risk$combined_risk_group <- 'Low Risk'
train_risk$combined_risk_group[combined_risk >= threshold] <- 'High Risk'

threshold
```

#Generating Kaplan Meier Curves for Combined Signature
```{r KM}
fit <- survfit(Surv(time, status) ~ combined_risk_group, data = train_risk)

ggsurvplot(
   fit,                     # survfit object with calculated statistics.
   data = train_risk,             # data used to fit survival curves.
   size = 1.2,
   palette = c("blue","red"),
   risk.table = FALSE,       # show risk table.
   legend = "top",
   conf.int = FALSE,
   pval = TRUE,             # show p-value of log-rank test..
   xlab = "Time in days" ,  # customize X axis label.
   ylab = "Disease-Free Survival Probability", # customize Y axis label.
   xlim = c(0,2500)
   #ylim = c(0.5,1)
   #cumcensor = TRUE      # plot the number of censored subjects at time t
 #  ncensor.plot.height = 0.25
)
```

##Validating the Signature

#Computing Cox regression coefficients for Test Set
```{r test_risk}

test_set <- lapply(1:length(varnames), function(i) {
  ifor <- coeffs[i]
  k <- (sprintf("%s", varnames[i]))
  feature_list <- test[,k]
  value11 <- feature_list*ifor
  df <- data.frame(value11)
})

store <- data.frame(test_set)
combined_risk <- as.data.frame(rowSums(store))
test_risk <- cbind(test,combined_risk)
colnames(test_risk)[colnames(test_risk) == "rowSums(store)"] <- "petul_risk"
```

#Validation Performance of Combined PetuL Risk
```{r test_cox}
quris.cox <- coxph(Surv(time, status) ~ IntuL_risk, data = test_risk)
summary(quris.cox)

#print hazard ratio
exp(quris.cox[["coefficients"]])
```

#Generating Risk Groups for Test Set
```{r risk_threshold}

test_risk$combined_risk_group <- 'Low Risk'
test_risk$combined_risk_group[combined_risk >= 1.03] <- 'High Risk'
```

```{r KM plot for training}
fit2 <- survfit(Surv(time, status) ~ combined_risk_group, data = test_risk)

ggsurvplot(
   fit2,                     # survfit object with calculated statistics.
   data = test_risk,             # data used to fit survival curves.
   size = 1.2,
   palette = c("red","blue"),
   risk.table = FALSE,       # show risk table.
   legend = "top",
   conf.int = FALSE,
   pval = TRUE,             # show p-value of log-rank test..
   xlab = "Time in days" ,  # customize X axis label.
   ylab = "Disease-Free Survival Probability", # customize Y axis label.
   xlim = c(0,2000),
   ylim = c(0,1)
   #cumcensor = TRUE      # plot the number of censored subjects at time t
 #  ncensor.plot.height = 0.25
)
```

## Validating the Prognostic Signature on the Validation Dataset

# Loading and tidying the clinical sheet

Loading the Emory clinical and imaging data into R.

```{r load, message=FALSE}
setwd("~/Documents/CCIPD/Thyroid")
Emory_peri <- read_excel("LabInvest_Supp.xlsx", sheet = 3)

Emory_peri <- Emory_peri %>%
  select(subj_id,gender,race,time,status,histologic_diagnosis,age_at_dx,t_stage,n_stage,m_stage,PetuL_index,PetuL_risk,IntuL_index,IntuL_risk)
```

# Cox regression model for validation set
```{r Cox_validation}
x1 <- Emory_peri[,c("PetuL_index","IntuL_index")]
x <- data.matrix(x1, rownames.force = NA)

y1 <- Emory_peri[,(4:5)] #time and status
y <- data.matrix(y1, rownames.force = NA)

#cox proportional hazards model
featureNames <- colnames(x)
varnames = sapply(1:length(featureNames), function(i){
  (paste0(featureNames[i]))
})

iformula <- as.formula(sprintf("Surv(time, status) ~ %s ", paste(varnames, collapse='+')))  
res.cox <- coxph(iformula, data = Emory_peri)
summary(res.cox)
```

# Creating a signature using Cox regression coefficients 
```{r generate_risk_scores}

coeffs <- res.cox[["coefficients"]] + 1
val_set <- lapply(1:length(varnames), function(i) {
  ifor <- coeffs[i]
  k <- (sprintf("%s", varnames[i]))
  feature_list <- Emory_peri[,k]
  value11 <- feature_list*ifor
  df <- data.frame(value11)
})

store <- data.frame(val_set)
combined_risk <- as.data.frame(rowSums(store))
Emory_risk <- cbind(Emory_peri,combined_risk)
colnames(Emory_risk)[colnames(Emory_risk) == "rowSums(store)"] <- "combined_risk"
```

# Cox Model Performance of PetuL Risk
```{r signature_perf}
quris.cox <- coxph(Surv(time, status) ~ PetuL_risk, data = Emory_risk)
summary(quris.cox)

#print hazard ratio
exp(quris.cox[["coefficients"]])
```

# Cox Model Performance of IntuL Risk
```{r signature_perf}
quris.cox <- coxph(Surv(time, status) ~ IntuL_risk, data = Emory_risk)
summary(quris.cox)

#print hazard ratio
exp(quris.cox[["coefficients"]])
```

# Cox Model Performance of Combined Risk
```{r signature_perf}
quris.cox <- coxph(Surv(time, status) ~ combined_risk, data = Emory_risk)
summary(quris.cox)

#print hazard ratio
exp(quris.cox[["coefficients"]])

Emory_risk_scores <- na.omit(rowSums(store))
Emory_threshold <- quantile(Emory_risk_scores, 0.7) # same threshold as training set
Emory_risk$combined_risk_group <- 'Low Risk'
Emory_risk$combined_risk_group[combined_risk >= Emory_threshold] <- 'High Risk'

fit_emory <- survfit(Surv(time, status) ~ combined_risk_group, data = Emory_risk)

ggsurvplot(
   fit_emory,                     # survfit object with calculated statistics.
   data = Emory_risk,             # data used to fit survival curves.
   size = 1.2,
   palette = c("red","blue"),
   risk.table = TRUE,       # show risk table.
   legend = "top",
   conf.int = FALSE,
   pval = TRUE,             # show p-value of log-rank test..
   xlab = "Time in days" ,  # customize X axis label.
   ylab = "Disease-Free Survival Probability", # customize Y axis label.
   xlim = c(0,2500),
   ylim = c(0,1)
   #cumcensor = TRUE      # plot the number of censored subjects at time t
 #  ncensor.plot.height = 0.25
)
```